Cross-scale integration of knowledge for predicting species ranges: a metamodeling framework

### 2014-10-14 tex/manuscript.tex
Finished first draft for submission

\title{Cross-scale integration of knowledge for predicting species ranges: a metamodeling framework}
\date{}
\author[1,2,13]{Matthew V. Talluto (mtalluto@gmail.com)}
\author[1,2]{Isabelle Boulangeat (isabelle.boulangeat@gmail.com)}
\author[3]{Aitor Ameztegui (ameztegui@gmail.com)}
\author[4]{Isabelle Aubin (Isabelle.Aubin@NRCan-RNCan.gc.ca)}
\author[1,2,5]{Dominique Berteaux (Dominique\_Berteaux@uqar.ca)}
\author[1,2]{Alyssa Butler (ca.butler10@gmail.com)}
\author[6,7]{Frédérik Doyon (Frederik.Doyon@uqo.ca)}
\author[8]{C. Ronnie Drever (cdrever@TNC.ORG)}
\author[9]{Marie-Josée Fortin (fortinmj@gmail.com)}
\author[1]{Tony Franceschini (Tony.Franceschini@uqar.ca)}
\author[10]{Jean Liénard (jean.lienard@gmail.com)}
\author[4]{Dan McKenney (Dan.McKenney@nrcan-rncan.gc.ca)}
\author[2,3]{Kevin A. Solarik (kevinsolarik@hotmail.com)}
\author[10]{Nickolay Strigul (nick.strigul@vancouver.wsu.edu)}
\author[11, 12]{Wilfried Thuiller (wilfried.thuiller@ujf-grenoble.fr)}
\author[1,2]{Dominique Gravel (dominique\_gravel@uqar.ca)}
\affil[1]{Département de biologie, Université du Québec à Rimouski, Rimouski, Quebec, Canada}
\affil[2]{Quebec Centre for Biodiversity Science, Montreal, Quebec, Canada}
\affil[3]{Centre d'Étude de la Forêt, Département des sciences biologiques, Université du Québec à Montréal, Montreal, Quebec, Canada}
\affil[4]{Great Lakes Forestry Centre, Canadian Forest Service, Natural Resources Canada, Sault Ste Marie, Ontario, Canada}
\affil[5]{Centre for Northern Studies, Université du Québec à Rimouski, Rimouski, Quebec, Canada}
\affil[6]{Université du Québec en Outaouais, Gatineau, Quebec, Canada}
\affil[7]{Institut des Sciences de la Forêt Tempérée (ISFORT), Ripon, Quebec, Canada}
\affil[8]{The Nature Conservancy Canada, Ottawa, Ontario, Canada}
\affil[9]{Department of Ecology and Evolutionary Biology, University of Toronto, Toronto, Ontario, Canada}
\affil[10]{Department of Mathematics, Washington State University Vancouver, Washington, USA}
\affil[11]{Université Grenoble Alpes, Laboratoire d’Ecologie Alpine (LECA), F-38000 Grenoble, France}
\affil[12]{CNRS, Laboratoire d’Ecologie Alpine (LECA), F-38000 Grenoble, France}
\affil[13]{Author for correspondance. Address: Departament de Biologie, chimie, et geographie, 300, Allée des Ursulines, Rimouski, Quebec G5L 3A1, Canada}

@@

	% abstract limited to 200 words
	% main text excludes abstract, acknowledgements, references, table and figure legends; limited to 7500 words
	% boxes limited to 750 words each
	% figure captions limited to 150 words each
	\textbf{Word counts: } Abstract: 200; Main Text: 6086; Boxes: 517

	\textbf{Number of references:} 79 % limited to 80

@@

\defcitealias{IPCC2001}{IPCC, 2001} % this is needed to make the organizational citation in the following paragraph work correctly
For the metamodel, we used a dataset of 1013 presence and 13863 absence records at 0.5 degree resolution.
We used a binomial GLM to relate the suitability for sugar maple to 3 climate variables: the number of degree days, mean summer precipitation, and the ratio of mean annual precipitation to potential evapotranspiration.
These variables were selected from the initial set of 6 variables from \citet{Morin2009}; see Appendix S2 for details on model construction and selection.
We selected a GLM for its simplicity and interpretability because our focus was on demonstrating the framework, but more complex methods (e.g., GAM) are compatible with the framework.
For projection, we used predictions for 2080 from the HadCM3 GCM \citep{Pope2000} driven by the A2 emission scenario \citep{Nakicenovic2000}, and used the parameters of the GLM to forecast suitability into the future (i.e., \(\psi_N\)).
Phenofit produces predictions at continental scales; thus the scaling function \(g\) was incorporated within Phenofit itself and we performed model integration directly on the outputs.
To perform the integration, we used a Metropolis-Hastings scheme within \ac{MCMC} to condition the parameters of the metamodel on two likelihood functions.
The first was the naive SDM, where the likelihood of a ``success'' in the presence/absence dataset was a Bernoulli density with the (logit-transformed) probability expressed as a linear function of the climate predictors.
For the sub-model, we considered that Phenofit, being a process-based model, may perform better when projecting into the future, particularly if suitability is primarily driven by the phenological match to the environment.
Thus, we used the predicted future probability of presence as the basis for generating simulated future presence-absence datasets.
This procedure inherently incorporates some uncertainty by randomly determining where the species is present (though uncertainty in the probabilities themselves could not be incorporated, as it is not possible to obtain them from Phenofit).
Thus, the sub-model likelihood is simply the probability, given the predictions of the metamodel, of observing a simulated dataset \(X_s\) that is randomly drawn from the phenofit predictions \(\psi_s\); i.e., \(X_s = f(\psi_s)\):
\begin{equation}
\label{eq:integrated2}
	p( \theta_M \mid X_M, D_M, \psi_s, \theta_S )
	\propto 
	p( X_S = f(\psi_s)\mid \theta_M, \theta_S )
	p( Y_M \mid \theta_M, D_M ) 
	p( \theta_M )
	p( \theta_S )
\end{equation}
where \(\theta_M\) is the metamodel, 
\(X_M\) is the vector of contemporary presences and absences, 
\(D_M\) is the matrix of climate predictors,
and \(\theta_S\) is the model relating the Phenofit predictions to the metamodel.
Furthermore, the integration is performed on the \emph{predictions} of the sub-model directly, rather than on compatible parameters as in Example 1. 
Thus, the metamodel must appear in the sub-model likelihood.

Considering the models individually, the predictions of Phenofit were much more conservative than those of the naive model (Fig \ref{fig:ex2}).
Although both models predicted a northward range shift, the change under Phenofit was limited to approximately 200 km north of the present range limit of the species, compared with more than 900 km for the naive model.
Phenofit also predicted little change in the southern range limit of the species, while the naive model projected a lack of suitability over the southern half of the present distribution.
Model integration produced predictions that were intermediate between the extremes set by the individual models (Fig \ref{fig:ex2}C).
The northern limit of the species was forecast to move by approximately 400--500 km, with reduced suitabilities but little extirpation throughout the southern portion of the present range.
Uncertainty in the naive model was relatively low, particularly for the southern portion of the range where the model predicted extirpation (Fig \ref{fig:ex2}D).
However, uncertainty in the predictions was locally extreme, particularly in portions of eastern Canada.
Prediction uncertainty was generally increased for the integrated model relative to the naive model (Fig \ref{fig:ex2}E), likely due to the large amount of disagreement in the predictions of the two models.
Unfortunately, validation of future projections is not possible due to a lack of data with which to compare model predictions.
However, model integration in this case provides a greater degree of humility in interpreting the predictions; for example, the naive model predicted with a high degree of certainty that much of the southern portion of the present range of sugar maple will be unsuitable by 2080.
The integrated model suggests that we should not be so confident in this interpretation.
Furthermore, the relatively high variance in predictions in the south suggests that further exploration of limiting factors in that region may be useful in providing further understanding and more precise predictions for changes to the distribution of this species.

@@

%TC:ignore
\nolinenumbers
\section*{Acknowledgements}
We acknowledge funding from the Quebec Centre for Biodiversity Science as well as NSERC strategic grant 430393-12. Comments from Charles Canham and William Godsoe improved a previous version of this manuscript. WT received support funding from the European Research Council under the European Community's Seven Framework Programme FP7/2007-2013 Grant Agreement no. 281422 (TEEMBIO). 
%TC:endignore

@@

>>>>

\title{Cross-scale integration of knowledge for predicting species ranges: a metamodeling framework}
\date{}
\author[1,2,13]{Matthew V. Talluto (mtalluto@gmail.com)}
\author[1,2]{Isabelle Boulangeat (isabelle.boulangeat@gmail.com)}
\author[3]{Aitor Ameztegui (ameztegui@gmail.com)}
\author[4]{Isabelle Aubin (Isabelle.Aubin@NRCan-RNCan.gc.ca)}
\author[1,2,5]{Dominique Berteaux (Dominique\_Berteaux@uqar.ca)}
\author[1,2]{Alyssa Butler (ca.butler10@gmail.com)}
\author[6,7]{Frédérik Doyon (Frederik.Doyon@uqo.ca)}
\author[8]{C. Ronnie Drever (cdrever@TNC.ORG)}
\author[9]{Marie-Josée Fortin (fortinmj@gmail.com)}
\author[1]{Tony Franceschini (Tony.Franceschini@uqar.ca)}
\author[10]{Jean Liénard (jean.lienard@gmail.com)}
\author[4]{Dan McKenney (Dan.McKenney@nrcan-rncan.gc.ca)}
\author[2,3]{Kevin A. Solarik (kevinsolarik@hotmail.com)}
\author[10]{Nikolay Strigul (nick.strigul@vancouver.wsu.edu)}
\author[11, 12]{Wilfried Thuiller (wilfried.thuiller@ujf-grenoble.fr)}
\author[1,2]{Dominique Gravel (dominique\_gravel@uqar.ca)}
\affil[1]{Département de biologie, Université du Québec à Rimouski, Rimouski, Quebec, Canada}
\affil[2]{Quebec Centre for Biodiversity Science, Montreal, Quebec, Canada}
\affil[3]{Centre d'Étude de la Forêt, Département des sciences biologiques, Université du Québec à Montréal, Montreal, Quebec, Canada}
\affil[4]{Great Lakes Forestry Centre, Canadian Forest Service, Natural Resources Canada, Sault Ste Marie, Ontario, Canada}
\affil[5]{Centre for Northern Studies, Université du Québec à Rimouski, Rimouski, Quebec, Canada}
\affil[6]{Université du Québec en Outaouais, Gatineau, Quebec, Canada}
\affil[7]{Institut des Sciences de la Forêt Tempérée (ISFORT), Ripon, Quebec, Canada}
\affil[8]{The Nature Conservancy Canada, Ottawa, Ontario, Canada}
\affil[9]{Department of Ecology and Evolutionary Biology, University of Toronto, Toronto, Ontario, Canada}
\affil[10]{Department of Mathematics, Washington State University, Vancouver, Washington, USA}
\affil[11]{Université Grenoble Alpes, Laboratoire d’Ecologie Alpine (LECA), F-38000 Grenoble, France}
\affil[12]{CNRS, Laboratoire d’Ecologie Alpine (LECA), F-38000 Grenoble, France}
\affil[13]{Author for correspondance. Address: Departament de Biologie, chimie, et geographie, 300, Allée des Ursulines, Rimouski, Quebec G5L 3A1, Canada}

@@

	% abstract limited to 200 words
	% main text excludes abstract, acknowledgements, references, table and figure legends; limited to 7500 words
	% boxes limited to 750 words each
	% figure captions limited to 150 words each
	\textbf{Word counts: } Abstract: 200; Main Text: 5953; Boxes: 516

	\textbf{Number of references:} 80 % limited to 80

@@

\defcitealias{IPCC2001}{IPCC, 2001} % this is needed to make the organizational citation in the following paragraph work correctly
For the metamodel, we used a dataset of 1013 presence and 13863 absence records at 0.5 degree resolution.
We used a binomial GLM to relate the suitability for sugar maple to 3 climate variables: the number of degree days, mean summer precipitation, and the ratio of annual precipitation to potential evapotranspiration.
These variables were selected from the initial set of 6 variables from \citet{Morin2009}; see Appendix S2 for details on model construction and selection.
We selected a GLM for its simplicity and interpretability because our focus was on demonstrating the framework, but more complex methods (e.g., GAM) are compatible with the framework.
For projection, we used predictions for 2080 from the HadCM3 GCM \citep{Pope2000} driven by the A2 emission scenario \citep{Nakicenovic2000}, and used the parameters of the GLM to forecast suitability into the future (i.e., \(\psi_N\)).
Phenofit produces predictions at continental scales; thus the scaling function \(g\) was incorporated within Phenofit itself and we performed model integration directly on the outputs.
To perform the integration, we used a Metropolis-Hastings scheme within \ac{MCMC} to condition the parameters of the metamodel on two likelihood functions.
The first was the naive SDM, where the likelihood of a ``success'' in the presence/absence dataset was a Bernoulli density with the (logit-transformed) probability expressed as a linear function of the climate predictors.
For the sub-model, we considered that Phenofit, being a process-based model, may perform better when projecting into the future, particularly if suitability is primarily driven by the phenological match to the environment.
Thus, we used the predicted future probability of presence as the basis for generating simulated future presence-absence datasets.
This procedure inherently incorporates some uncertainty by randomly determining where the species is present (though uncertainty in the probabilities themselves could not be incorporated, as it is not possible to obtain them from Phenofit).
Thus, the sub-model likelihood is simply the probability, given the predictions of the metamodel, of observing a simulated dataset \(X_s\) that is randomly drawn from the phenofit predictions \(\psi_s\); i.e., \(X_s = f(\psi_s)\):
\begin{equation}
\label{eq:integrated2}
	p( \theta_M \mid X_M, D_M, \psi_s, \theta_S )
	\propto 
	p( X_S = f(\psi_s)\mid \theta_M, \theta_S )
	p( Y_M \mid \theta_M, D_M ) 
	p( \theta_M )
	p( \theta_S )
\end{equation}
where \(\theta_M\) is the metamodel, 
\(X_M\) is the vector of contemporary presences and absences, 
\(D_M\) is the matrix of climate predictors,
and \(\theta_S\) is the model relating the Phenofit predictions to the metamodel.
Furthermore, the integration is performed on the \emph{predictions} of the sub-model directly, rather than on compatible parameters as in Example 1. 
Thus, the metamodel must appear in the sub-model likelihood.

Considering the models individually, the predictions of Phenofit were much more conservative than those of the naive model (Fig \ref{fig:ex2}).
Although both models predicted a northward range shift, the change under Phenofit was limited to approximately 200 km north of the present range limit of the species, compared with more than 900 km for the naive model.
Phenofit also predicted little change in the southern range limit of the species, while the naive model projected a lack of suitability over the southern half of the present distribution.
Model integration produced predictions that were intermediate between the extremes set by the individual models (Fig \ref{fig:ex2}C).
The northern limit of the species was forecast to move by approximately 400--500 km, with reduced suitabilities but little extirpation throughout the southern portion of the present range.
Uncertainty in the naive model was relatively low, particularly for the southern portion of the range where the model predicted extirpation (Fig \ref{fig:ex2}D).
However, uncertainty in the predictions was locally extreme, particularly in portions of eastern Canada.
Prediction uncertainty was generally increased for the integrated model relative to the naive model (Fig \ref{fig:ex2}E), likely due to the large amount of disagreement in the predictions of the two models.

Although validation is an important tool for analyzing the performance of SDMs, it is not possible to validation predictions for the future due to a lack of data with which to compare model predictions.
However, model integration in this case provides a greater degree of humility in interpreting the predictions; for example, the naive model predicted with a high degree of certainty that much of the southern portion of the present range of sugar maple will be unsuitable by 2080.
The integrated model suggests that we should not be so confident in this interpretation.
The naive model predicts suitability based on the values of selected climate variables (in this case, number of degree days, mean summer precipitation, and the ratio of annual precipitation to potential evapotranspiration).
These variables were selected because they provided the best fit to the present distribution; however, this does not mean these variables are the most important climatic drivers of the processes generating the distribution, nor that they will continue to be important in the future.
In contrast, Phenofit predicts fitness based on how climate affects phenological timings, frost injury, reproduction, and survival \citep{Chuine2001, Morin2009}.
Thus, climatic factors that ultimately limit species distribution might be quite different under Phenofit, as illustrated by the substantial differences in the potential future distributions predicted by Phenofit and the naive model.
In the absence of information about which model is ``best,'' the integrated model provides a consensus view of these two very different models by fitting parameters that maximize the joint likelihood of the presence-absence dataset and the Phenofit predictions.
Furthermore, the integrated model provides robust estimates of uncertainty; in the southern portion of the range, the relatively high variance in predictions reflects the disagreement in the models and suggests that further exploration of limiting factors in that region may be useful in improving understanding and providing more precise predictions for changes to the distribution of sugar maple.

@@

%TC:ignore
\nolinenumbers
\section*{Acknowledgements}
We acknowledge funding from the Quebec Centre for Biodiversity Science as well as NSERC strategic grant 430393-12. Comments from Charles Canham, William Godsoe, and Tamara Münkemüller improved a previous version of this manuscript. WT received support funding from the European Research Council under the European Community's Seven Framework Programme FP7/2007-2013 Grant Agreement no. 281422 (TEEMBIO). 
%TC:endignore

@@

\section*{Supporting Information}
Additional Supporting Information may be downloaded via the online version of this article at Wiley Online Library (www.ecologyletters.com). As a service to our authors and readers, this journal provides supporting information supplied by the authors. Such materials are peer-reviewed and may be reorganized for online delivery, but are not copy-edited or typeset. Technical support issues arising from supporting information (other than missing files) should be addressed to the authors.
